MS_Imp_Guidelines

### 2023-10-16 Markdown/Manuscript/Child_RMDs/MS_Conclusion.Rmd
close to submission update - except tables

---
title: "MS_Conclusion"
author: "Author"
date: "`r Sys.Date()`"
output: html_document
---
`r ifelse(is.na(params$X@conclusion_1L_chr),"",params$X@conclusion_1L_chr)` 
<!-- Standards generalisable, Toolkit less os -->

>>>>

---
title: "MS_Conclusion"
author: "Author"
date: "`r Sys.Date()`"
output: html_document
---
`r ifelse(is.na(params$X@conclusion_1L_chr),"",params$X@conclusion_1L_chr)` 
We have identified criteria that can be used to systematically assess the extent to which the computational implementation of health economic models adheres to transparency, reusability and updatability criteria. We have developed a prototype software framework that can support the implementation of CHEMs that largely meet these criteria. Our framework can be used as a prototype for developing future software frameworks to support ethical implementation of CHEMs.

### 2023-10-16 Markdown/Manuscript/Child_RMDs/MS_Discussion.Rmd
close to submission update - except tables

<!-- Talk about DARTH (https://doi.org/10.1007/s40273-019-00837-x )/ BCEA -->
Ethical practice is a core expectation of health researchers and computational methods underpin most quantitative research, yet an understanding of what constitutes ethical computational modelling practice in health is underdeveloped [@10.3389/fpubh.2017.00068]. The modeller responsibilities, enabling model attributes and model implementation assessment criteria that we propose can help address this gap.

The ethical responsibilities and enabling model attributes we describe have both commonalities and distinctive features compared to a previous ethical framework for computational modelling in public health [@10.3389/fpubh.2017.00068]. The authors of that framework propose 13 questions to evaluate ethical risk across the four criteria of independence, transparency, beneficence and justice. Their descriptions of the four criteria at least partially map to either our proposed modeler responsibilities (“justice” to “social acceptability”, “independence” to “adequacy for purpose” and “beneficence” to “beneficial impact”) or enabling model attributes (“transparency”). However, while our six assessment criteria are specific to three attributes (TRU) of the computational implementation of the model, the prior ethical framework includes questions relevant to the conceptual and mathematical models and the potential impacts of model use. Examples of these more general evaluation questions include (for the justice criterion) “is any lack of knowledge about important parameters attributable to uncertainty or variability?” and (for the beneficence criterion) “if a policy is based on the model evidence, is it more likely to be effective and beneficial than a decision made in the absence of the model?”. The less numerous and more focused assessment criteria we propose may potentially be more tractable to implement in reviews of models authored by third parties and as the basis for designing software frameworks to support ethical computational model implementation.

Currently, many if not most existing CHEMs are insufficiently transparent [@Jalali2021; @McManus2019; @Bermejo2017; @Ghabri2019], reusable [@Feenstra2022; @Emerson2019] and updatable [@Sampson2017; @10.3389/fpubh.2022.899874]]. <!-- DELETION --> Existing incentive structures for health economists generally do not promote facilitating peers to reuse their work. Currently, it can take “an extraordinary amount of idealism” to commit to authoring and maintaining research software [@anzt2020environment].

Reducing waste in research is a core responsibility of research funders [@chalmers2014increase] and funding the development of CHEMs that are not adequately understood, reused or updated is wasteful. Previously recommended strategies for more beneficial health economic research investments include support for harmonized ethical standards for model development [@10.3389/fpubh.2017.00068], methodological innovation to improve model transferability  [@craig2018taking], networks of modellers working on common health conditions [@Sampson2019], and  centralized infrastructure such as open source model repositories [@Pouwels2022] and a standard platform for model implementations [@Ghabri2019]. Development of software frameworks to support ethical CHEM implementations could enable and enhance each of these strategies.

As illustrated by Table \@ref(tab:checktb), we have developed a software framework that can help us to author a youth mental health model that largely satisfy our TRU criteria. However, we believe our software framework is currently too fragile to be anything more than a prototype for supporting the development needs other modelling teams and projects. A major reason for this distinction is that our software framework was developed with the needs of only one group of developers in mind – ourselves. We currently lack the resources required to adequately implement strategies to target factors such as user enjoyment, usability, active user-community and supporting resources that influence adoption of software frameworks [@myllarniemi2018development].

Our prototype framework has a number of features that subsequent work to develop ethical software frameworks may find useful to incorporate. Firstly, developing a software framework to work within an existing and widely used open source programming language such as R or python, can keep framework scope relatively narrow (making it more tractable to develop, maintain and learn) while readily leveraging and coherently integrate other modelling tools written in that language (e.g. the dependency libraries we list in Table \@ref(tab:cpkgs)). Secondly, implementation that combines both object oriented and functional programming paradigms can avail of the modular and syntactical simplicity benefits of the former, while limiting needless bundling of code artefacts. Thirdly, a sensible trade-off needs to be found between transparent code implementation (which requires clear and sufficiently detailed documentation) and Agile Software Development (for which a foundational principle is prioritizing the development of working code over writing documentation [@beck2001manifesto]). Our software framework makes this trade off by enforcing the use of consistent code naming conventions and file organisation which in turn enables automated generation of simple documentation at every code update. All model data-structures and algorithms are therefore always documented (at least minimally, with machine authored content), meaning model developers have a requirement to write customized documentation less frequently.

>>>>

<!-- Talk about DARTH (https://doi.org/10.1007/s40273-019-00837-x )/ BCEA -->
Ethical practice is a core expectation of health researchers and computational methods underpin most quantitative research, yet an understanding of what constitutes ethical computational modelling practice in health is underdeveloped [@10.3389/fpubh.2017.00068]. We have addressed this gap with our proposed modeller and funder responsibilities, enabling model attributes and model implementation assessment criteria.

<!-- The less numerous and more focused assessment criteria we propose may potentially be more tractable to implement in reviews of models authored by third parties and as the basis for designing software frameworks to support ethical computational model implementation. -->

<!-- Two issues for funders - allow support for beneficial use, foster viable communities. -->

Currently, many if not most existing CHEMs are insufficiently transparent [@Jalali2021; @McManus2019; @Bermejo2017; @Ghabri2019], not reusable [@Feenstra2022; @Emerson2019], nor updatable [@Sampson2017; @10.3389/fpubh.2022.899874]. <!-- DELETION --> Addressing these issues will require changes to the tools, practices and funding underpinning health economic model development.

As illustrated by Table \@ref(tab:checktb), we have developed a prototype software framework that can be used to implement youth mental health CHEMs that largely satisfy our TRU criteria. However, in its current form our software framework is too fragile to be anything more than a prototype for supporting the development needs other modelling teams and projects. We developed our prototype software framework was developed with the needs of only one group of developers in mind – ourselves. We currently lack the resources required to adequately implement strategies to target factors such as user enjoyment, usability, active user-community and supporting resources that influence adoption of software frameworks [@myllarniemi2018development].

Our prototype has a number of features that subsequent work to develop CHEM software frameworks may find useful to incorporate. Firstly, developing a software framework to work within an existing and widely used open source programming language such as R or python, can keep framework scope relatively narrow. This makes it more tractable to develop, maintain and learn, and can integrate other modelling tools written in that language (e.g., the dependency libraries we list in Table \@ref(tab:cpkgs)). Secondly, implementation that combines both object oriented and functional programming paradigms can avail of the modular and syntactical simplicity benefits of the former, while limiting needless bundling of code artefacts. Thirdly, a sensible trade-off needs to be found between transparent code implementation (which requires clear and sufficiently detailed documentation) and Agile Software Development (for which a foundational principle is prioritising the development of working code over writing documentation [@beck2001manifesto]). Our software framework makes this trade off by enforcing the use of consistent code naming conventions and file organisation which in turn enables automated generation of simple documentation at every code update. All model data-structures and algorithms are therefore always documented (at least minimally, with machine authored content), meaning model developers have a requirement to write customised documentation less frequently.

Reducing waste in research is a core responsibility of research funders [@chalmers2014increase] and funding the development of CHEMs that are not adequately understood, reused or updated is wasteful. Existing incentive structures for health economists generally do not promote facilitating peers to reuse their work and funding is rarely adequate to maintain an routinely update a model. Currently, it can take “an extraordinary amount of idealism” to commit to authoring and maintaining research software [@anzt2020environment]. Previously recommended strategies for more beneficial health economic research investments include support for harmonised ethical standards for model development [@10.3389/fpubh.2017.00068], methodological innovation to improve model transferability [@craig2018taking], networks of modellers working on common health conditions [@Sampson2019], and  centralized infrastructure such as open source model repositories [@Pouwels2022] and a standard platform for model implementations [@Ghabri2019]. Development of software frameworks to support ethical CHEM implementations could enable and enhance each of these strategies.

### 2023-10-16 Markdown/Manuscript/Child_RMDs/MS_Imp_Example.Rmd
close to submission update - except tables

##	Economic topics
Currently, we are using the ready4 software framework to develop, apply and share youth mental health CHEMs in four of the twelve domains of health economics identified by Wagstaff and Culyer [@wagstaff2012four]:

-	health and its value (our projects: utility mapping models);

@@

Potential future directions are to supplement this work with CHEMs in two additional Wagstaff and Culyer domains of public health (to model the impact of selected fiscal policy and regulation options on young people's mental health) and human resources (to model the supply and behaviours of the youth mental health workforce). Our ultimate aim is to flexibly combining all our CHEMs in analyses that help answer questions in two additional Wagstaff and Culyer domains:

@@

Although principally interested in using our CHEMs to answer policy questions relating to the mental health of young people in Australia, we want to facilitate CHEM transferability to other jurisdictions. Our CHEMs are therefore being derived from and applied to real data (which can be assumption, empirical or simulated, so long as it is appropriate for use in analysis intended to inform decision-making) from Australia. Additionally, we plan for all our CHEMs to be distributed with toy (fake) data to help demonstrate the potential use of these CHEMS in other decision contexts.

## Case study: health and its value
We have previously described [@Hamilton2021.07.07.21260129] a study to develop utility mapping models for use in samples of young people presenting to primary mental health services. The ready4 software framework was used in that study to develop CHEM modules, supply those modules with data and implement modelling analyses, creating the following artefacts:

@@

We created a checklist (Table \@ref(tab:checktb)) that we used to subjectively assess these study outputs against TRU criteria. For each criterion, we provided a global assessment of whether it was met using the responses “yes”, “no” or “partial”.  We believe the outputs from our utility mapping study may be assessable as having satisfactorily met five of the six criteria (T1, T2, R1, R2 and U1) and to have partially met one criterion (U2). The main shortcomings that we identified when applying the assessment criteria was that we have yet to adequately implement unit testing of the R libraries we authored as part of this study.

>>>>

##	Economic topics
Currently, we are using the ready4 software framework to develop, apply, and share youth mental health CHEMs in four of the twelve domains of health economics identified by Wagstaff and Culyer [@wagstaff2012four]:

-	health and its value (our projects: models to map psychological and functional measures to health utility);

@@

Potential future directions are to supplement this work with CHEMs in two additional Wagstaff and Culyer domains of: (i) public health (to model the impact of selected fiscal policy and regulation options on young people's mental health); and (ii) human resources (to model the supply and behaviours of the youth mental health workforce). Our ultimate aim is to flexibly combining all our CHEMs in analyses that help answer questions in two additional Wagstaff and Culyer domains:

@@

Although we are principally interested in using CHEMs to answer policy questions relating to the mental health of young people in Australia, we want to facilitate CHEM transferability to other jurisdictions. Our CHEMs are being derived from and applied to real data (which can be assumption, empirical or simulated, so long as it is appropriate for use in analysis intended to inform decision-making) from Australia. To help demonstrate the potential use of CHEMs in other decision contexts, we also create toy datasets. Data created for illustration purposes is prominently labelled as not for use in decision-analysis.

## Case study: health and its value
We have previously describeda study [@Hamilton2021.07.07.21260129] to develop utility mapping models for use in samples of young people presenting to primary mental health services. The ready4 software framework was used in that study to develop CHEM modules, supply those modules with data and implement modelling analyses, creating the following artefacts:

@@

We created a checklist (Table \@ref(tab:checktb)) that we used to subjectively assess these study outputs against TRU criteria. For each criterion, we provided a global assessment of whether it was met using the responses “yes”, “no” or “partial”.  We believe our utility mapping study CHEMs have satisfactorily met five of the six criteria (T1, T2, R1, R2 and U1) and have partially met one criterion (U2). The main shortcomings that we identified when applying the assessment criteria was that we have yet to adequately implement unit testing of the R libraries authored as part of this study.

### 2023-10-16 Markdown/Manuscript/Child_RMDs/MS_Imp_Guidelines.Rmd
close to submission update - except tables

We considered prior literature on modelling practice and our own professional experience to identify: (i) some core ethical responsibilities of CHEM developers; (ii) attributes of CHEMs that can suggest fulfillment of these responsibilities; and (iii) criteria against which these attributes can be assessed.

## Ethical responsibilities of CHEM developers 
For simplicity, we sought to identify one overarching ethical responsibility for modellers at each distinct phase of a CHEM project. As CHEMs are implemented as software, software development lifecycle (SDLC) models may provide an appropriate basis for identifying the main phases of a CHEM project. There are multiple SDLC models, each taking a different approach to the number, naming, definition, sequencing and iteration of project phases [@ruparelia2010software]. However, components of most SDLC models can broadly be mapped to the concepts of *planning* (in which software requirements and design specification are agreed), *development* (in which software is authored and tested) and *deployment* (in which software is made available as production ready tools for end-users). We contend that ethical responsibilities of computational modellers identified in prior literature [@10.3389/fpubh.2017.00068; @pliakos2021ethics; @thompson2022escape; @calder2018computational] can be summarised under the following three overarching goals (one for each phase):

- **make socially acceptable value judgments** (when planning CHEM projects);

- **ensure model fitness for purpose** (when developing CHEMs); and

- **promote socially beneficial use of models** (once CHEMs have been deployed).

There is significant scope for health economists to improve fulfilment of each of these responsibilities.

Misalignment between the values of computational model developers and those of the population groups affected by decisions based on their models presents significant ethical risks [@thompson2022escape; @duckett2022journey]. The value judgments of CHEM developers influence the assumptions, selection of model features and standards for evidence that shape the model development process [@HARVARD2020112975]. These value judgments are rarely made explicit, omissions that may lead to socially unacceptable policy recommendations [@duckett2022journey]. Ideally, it should be straightforward for CHEMs to be modified to reflect alternative value judgments. However, such adaptability depends in part on the approach taken to computational model implementation (see Table \@ref(tab:proscons)).

@@

Computational modellers have duties both to take sufficient care that a computational model is adequate for the explicit purpose for which it was developed and to provide potential third-party users with the means of assessing its adequacy for their proposed purposes [@Erdemir2020; @Feenstra2022; @thompson2019escape; @10.3389/fpubh.2017.00068]. However, it is common for health economic models to have serious methodological flaws [@carletto_zanuzzi_sammarco_russo_2020; @WONDER2015467]; insufficient validation [@Ghabri2019; @kolovos2017model; @haji2013model], poor reproducibility [@Jalali2021; @McManus2019; @Bermejo2017]; and undeclared errors [@Radeva2020]. Appropriate computational implementation choices can help address many of these shortcomings, for example by automating quality assurance checks and facilitating manual reviews by third parties.

A health economic model will have limited public benefit if it not much used, if it is mis-used or when its acceptability and adequacy rapidly decay. Reuse of CHEMs as components of other models can potentially make model development more efficient [@Arnold2010; @garcia2021cost]. However, health economic models face challenges related to transferability across jurisdictions [@garcia2021cost] that create barriers to reuse. Without ongoing maintenance, a CHEM risks becoming less reliable with time [@garcia2021cost] and is at risk of being deployed for purposes for which it is poorly suited [@calder2018computational]. Currently, health economic models are rarely implemented computationally in a manner that facilitates routine updates [@Sampson2017], thus limiting the temporal window within which a CHEM can be validly applied. 
<!-- Reuse of models as components of other models can also make model implementation more efficient [@Arnold2010; @garcia2021cost] but the reusability of models is significantly shaped by computational implementation choices such as how model code and data are authored, documented, licensed and disseminated.  -->

@@

## CHEM attributes associated with ethical modelling practice
Health economist responsibilities during the planning, implementation and use of CHEMs are easier to state than to measure.  However, aspects of ethical modelling practice may be inferred from measurable attributes of CHEMs. As described in Table \@ref(tab:timelygls),  we believe that the creators of transparent, reusable and updatable (TRU) CHEMs are likely to have fulfilled a number of their ethical obligations.  We therefore selected these model attributes to use as the basis for deriving ethical assessment criteria.

@@

Taking steps to make computational models **transparent** is an important part of ethical public health modelling practice [@10.3389/fpubh.2017.00068]. Guidance on transparency in health economic modelling recommended that model code and data should be clearly documented [@Eddy2012]. Notably, the same guidelines, published over ten years ago, did not recommend sharing model code and data. However, more recent healthcare modelling guidance does recommend publicly dissemination of such artefacts [@Erdemir2020]. Online repository services such as Zenodo [@Zenodo2013] and Dataverse [@Dataverse2007] provide persistent storage solutions that generate a Digital Object Identifier (DOI) for each code and data collection. Ensuring that calculations are correct and consistent with model specifications is an essential part of CHEM quality assurance [@techver2019]. The extensiveness of such verification checks can be reported using the concept of code coverage [@ERICWONG2010188] - the proportion of model code that has been explicitly tested. Tests should ideally combine both unit tests (to verify that small, isolated sections of code produce the correct output when run independently) and acceptance tests (to verify that the correct output is produced when multiple code components are run together to perform tasks that meet core user-requirements [@martin2003agile]). The nature and extent of individual model authorship contributions can become unclear when models are implemented over longer time-frames with a large and changing group of collaborators [@thompson2022escape]. This issue can be addressed by use of online repository services such as GitHub [@github2007], that provide citation tools and can transparently record all individual code contributions to a modelling project over its lifecycle.

@@

Models should be **updatable** so that they remain valid for longer, evolving as new evidence emerges and the systems being modelled change [@Jenkins2021; @garcia2021cost]. Ensuring that a model is regularly reviewed to identify and implement required improvements is a recommended defense against model validity decay [@calder2018computational]. Sustainable maintenance of open source research software requires both a core development team and an active user community [@info:doi/10.2196/20028]. Online communities can be an efficient means of engaging model users in testing each version of a model, identifying issues and suggesting improvements. Services such as GitHub [@github2007] provide collaborative code development tools [@MERGEL2015464] that help elicit, integrate and reconcile contributions from multiple contributors and to ensure each update is uniquely identifiable and retrievable. It is also important that verification checks are rerun with each model update, a task that can be automated using the software development practice of continuous integration [@CI2017].  The risk of model revisions having unintended consequences for third party users can be mitigated through the use of deprecation conventions [@zhou2016api] that take an informative and staged approach to retiring outdated model code and data.
<!-- Currently, the core development team for a CHEM will be typically be funded to produce a project end-point deliverable whose specifications are well defined early in the project. For more complex and multi-purpose CHEMs, particularly those designed to be incorporated into decision support systems, it may be better for development teams to adopt Agile Software Development [@beck2001manifesto], an approach that has been recommended for complex public health software projects [@10.3389/fpubh.2022.899874]. An Agile model will be less clearly specified in the initial project plan, but will instead continually develop in response to the requirements and feedback of users, who are provided with an initial, simplified working version of the model at the earliest feasible opportunity. -->

@@

-	**T1**: All model code, non-confidential data and testing procedures and outcomes are available in open access repositories.

-	**T2**: It is easy to see who developed and tested each part of the CHEM and to identify the modelling team’s assumptions, judgments and theories about CHEM development and use.

@@

-	**R1**: Model code and data are implemented to facilitate both generalisability and transferability.

-	**R2**: Terms of use allow anyone to reuse model code and non-confidential data, in whole or in part, without charge, and for purposes that include the creation of derivative works.

@@

-	**U1**: Technical infrastructure is in place to support version control and collaboration with model users in the maintenance of a model.

-	**U2**: Each new release of a model is retested, with changes implemented to minimize disruptions for existing model users.

>>>>

We considered prior literature on modelling practice and our own professional experience to identify: (i) some core ethical responsibilities of CHEM developers and funders; (ii) attributes of CHEMs that can suggest fulfillment of these responsibilities; and (iii) criteria against which these attributes can be assessed.

## Ethical responsibilities of CHEM developers and funders
<!-- The ethical responsibilities and enabling model attributes we describe have both commonalities and distinctive features compared to a previous  -->
Boden and McKendrick [@10.3389/fpubh.2017.00068] proposed a framework for ethical computational modelling in public health based on the four ethical criteria of justice (concerning the social obligations of modellers to consider and communicate ethical issues), independence (concerning how modeller subjectivity shapes model design), transparency (concerning the need for policymakers to reliably evaluate model strengths and weaknesses) and beneficence / non-maleficence (concerning model quality and utility). However, the framework's authors mostly discuss these criteria in the context of a modelling project as a whole, including a model's conceptual and mathematical representations. The authors propose 13 questions to evaluate ethical risk across the four criteria, but of these only "is the model code open source or available on request" (for transparency) and "has the model been verified, i.e., does it do what the modeller wants it to do?" (for beneficence / non-maleficence) are specific to model's representation as software.

To reconceptualise these four criteria as ethical responsibilities that are specific to CHEM developers and funders, we first considered the distinct phases in a software development project. Software development lifecycle (SDLC) models take different approaches to the inclusion, naming, definition, sequencing and iteration of project phases [@ruparelia2010software]. However, SDLC models typically have components that can principally map to the concepts of *planning* (e.g. identification of system / user requirements and design specifications), *implementation* (e.g. development and testing) and *release* (e.g.deployment, system integration and maintenance). For simplicity, we propose one overarching ethical responsibility for CHEM developers and funders for each of these phases. These responsibilities are to:

(i) **make socially acceptable system requirements and design decisions** (during CHEM planning);

(ii) **ensure fitness for purpose** (of CHEM implementation); and

(iii) **support socially beneficial use** (on CHEM release).

Each responsibility combines different aspects of Boden and McKendrick's ethical criteria: responsibility (i) incorporates justice and independence; responsibility (ii) incorporates independence and transparency; and responsibility (iii) incorporates beneficence / non-maleficence and justice. Model developers and funders share these responsibilities as each one has implications for both professional practice and project resourcing.

<!-- . The authors of that framework propose 13 questions to evaluate ethical risk across the four criteria of  -->
<!-- . However, while our six assessment criteria are specific to three attributes (TRU) of the computational implementation of the model, the prior ethical framework includes questions relevant to the conceptual and mathematical models and the potential impacts of model use. Examples of these more general evaluation questions include (for the justice criterion) “is any lack of knowledge about important parameters attributable to uncertainty or variability?” and (for the beneficence criterion) “if a policy is based on the model evidence, is it more likely to be effective and beneficial than a decision made in the absence of the model?”.  -->

<!-- Multiple ethical considerations of computational modeling have previously been identified [@10.3389/fpubh.2017.00068; @pliakos2021ethics; @thompson2022escape; @calder2018computational],  -->

Misalignment between the values of computational model developers and those of the population groups affected by decisions based on their models presents significant ethical risks [@thompson2022escape; @duckett2022journey]. The value judgments of health economists influence the assumptions, selection of model features and standards for evidence that shape health economic model projects [@HARVARD2020112975]. These value judgments are rarely made explicit, omissions that may lead to socially unacceptable policy recommendations [@duckett2022journey]. modeller value judgments also determine the computational implementation of health economic models. For example, as illustrated in Table \@ref(tab:proscons), project resource constraints may require trade offs between desired CHEM **system requirements and design** features. These trade offs in turn determine how straightforward it is for CHEMs to be amended by third parties to reflect alternative value judgments in model conceptualisation and mathematical formalisation (Table \@ref(tab:timelygls)).

@@

Computational modellers have duties both to ensure computational model **fitness for purpose** and to provide potential third-party users with the means of assessing its adequacy for their proposed purposes [@Erdemir2020; @Feenstra2022; @thompson2019escape; @10.3389/fpubh.2017.00068]. However, it is common for health economic models to have serious methodological flaws [@carletto_zanuzzi_sammarco_russo_2020; @WONDER2015467]; insufficient validation [@Ghabri2019; @kolovos2017model; @haji2013model], poor reproducibility [@Jalali2021; @McManus2019; @Bermejo2017]; and undeclared errors [@Radeva2020]. Appropriate computational implementation choices can help address many of these shortcomings, for example by automating quality assurance checks and facilitating manual reviews by third parties.

A CHEM will have limited **social benefit** if it is rarely used, if mis-used or when its acceptability and adequacy rapidly decay. Reuse of CHEMs as components of other models can potentially make model development more efficient [@Arnold2010; @garcia2021cost]. However, health economic models face challenges related to transferability across jurisdictions [@garcia2021cost] that create barriers to reuse. Without ongoing maintenance, a CHEM risks becoming less reliable with time [@garcia2021cost] and is at risk of being deployed for purposes for which it is poorly suited [@calder2018computational]. Currently, health economic models are rarely implemented computationally in a manner that facilitates routine updates [@Sampson2017], thus limiting the temporal window within which a CHEM can be validly applied. 
<!-- Reuse of models as components of other models can also make model implementation more efficient [@Arnold2010; @garcia2021cost] but the reusability of models is significantly shaped by computational implementation choices such as how model code and data are authored, documented, licensed and disseminated.  -->

@@

## CHEM attributes associated with ethical modelling practice
The responsibilities of model developers and funders over a CHEM lifecycle are easier to state than to measure.  However, aspects of ethical modelling practice may be inferred from measurable attributes of CHEMs. As described in Table \@ref(tab:timelygls),  we believe that the creators of transparent, reusable and updatable (TRU) CHEMs are likely to have fulfilled a number of their ethical obligations.  We therefore selected these model attributes to use as the basis for deriving ethical assessment criteria.

@@

Taking steps to make computational models **transparent** is an important part of ethical public health modelling practice [@10.3389/fpubh.2017.00068]. Guidance on transparency in health economic modelling recommended that model code and data should be clearly documented [@Eddy2012]. Notably, the same guidelines, published over ten years ago, did not recommend sharing model code and data. However, more recent healthcare modelling guidance does recommend public dissemination of such artefacts [@Erdemir2020]. Online repository services such as Zenodo [@Zenodo2013] and Dataverse [@Dataverse2007] provide persistent storage solutions that generate a Digital Object Identifier (DOI) for each code and data collection. Ensuring that calculations are correct and consistent with model specifications is an essential part of CHEM quality assurance [@techver2019]. The extensiveness of such verification checks can be reported using the concept of code coverage [@ERICWONG2010188] - the proportion of model code that has been explicitly tested. Tests should ideally combine both unit tests (to verify that small, isolated sections of code produce the correct output when run independently) and acceptance tests (to verify that the correct output is produced when multiple code components are run together to perform tasks that meet core user-requirements [@martin2003agile]). The nature and extent of individual model authorship contributions can become unclear when models are implemented over longer time-frames with a large and changing group of collaborators [@thompson2022escape]. This issue can be addressed by use of online repository services such as GitHub [@github2007], that provide citation tools and can transparently record all individual code contributions to a modelling project over its lifecycle.

@@

Models should be **updatable** so that they remain valid for longer, evolving as new evidence emerges and the systems being modelled change [@Jenkins2021; @garcia2021cost]. Ensuring that a model is regularly reviewed to identify and implement required improvements is a recommended defence against model validity decay [@calder2018computational]. Sustainable maintenance of open source research software requires both a core development team and an active user community [@info:doi/10.2196/20028]. Online communities can be an efficient means of engaging model users in testing each version of a model, identifying issues and suggesting improvements. Services such as GitHub [@github2007] provide collaborative code development tools [@MERGEL2015464] that help elicit, integrate and reconcile contributions from multiple contributors and to ensure each update is uniquely identifiable and retrievable. It is also important that verification checks are rerun with each model update, a task that can be automated using the software development practice of continuous integration [@CI2017].  The risk of model revisions having unintended consequences for third party users can be mitigated through the use of deprecation conventions [@zhou2016api] that take an informative and staged approach to retiring outdated model code and data.
<!-- Currently, the core development team for a CHEM will be typically be funded to produce a project end-point deliverable whose specifications are well defined early in the project. For more complex and multi-purpose CHEMs, particularly those designed to be incorporated into decision support systems, it may be better for development teams to adopt Agile Software Development [@beck2001manifesto], an approach that has been recommended for complex public health software projects [@10.3389/fpubh.2022.899874]. An Agile model will be less clearly specified in the initial project plan, but will instead continually develop in response to the requirements and feedback of users, who are provided with an initial, simplified working version of the model at the earliest feasible opportunity. -->

@@

-	**T1**: Software files (executable code, testing procedures and outcomes and non-confidential data) are available in open access repositories.

-	**T2**: It is easy to see who funded, developed and tested each part of the CHEM and to identify the modelling team’s assumptions, judgments and theories about CHEM development and use.

@@

-	**R1**: Programming practices facilitate both generalisability and transferability.

-	**R2**: Terms of use place only ethical restrictions on reuse and allow anyone to ethically reuse model code and non-confidential data, in whole or in part, without charge, and for purposes that include the creation of derivative works.

@@

-	**U1**: Maintenance infrastructure is in place to support version control and collaboration with model users.

-	**U2**: Each new release of a model is systematically retested, with changes implemented to minimise disruptions for existing model users.

### 2023-10-16 Markdown/Manuscript/Child_RMDs/MS_Imp_Toolkit.Rmd
close to submission update - except tables

## R libraries 
A library in the R language will typically depend on multiple other R libraries, all of which potentially having different authors. As the number of third-party dependencies of an R library grows so does the fragility of that library (e.g. the library may cease to work as intended due to changes in one of its dependency libraries). To reduce the fragility of our framework we implemented it as multiple R libraries rather than one R library. In total we authored six novel R libraries to implement the ready4 framework, all of which have distinct purposes and dependencies (Table \@ref(tab:cpkgs)).

One framework library provides a **foundation for modular CHEM implementations**. In modular model implementations, modules need to be able to share inputs and outputs with each other and to be run as independent models [@barros2023empowering]. To achieve this, the foundation framework library defines a template CHEM module (using R’s S4 class system), which can be used to create other CHEM modules with a common set of inherited properties. One of these inherited properties is a novel syntax of 15 core commands that enable CHEM module algorithms to be consistently named. The foundation library also contains tools for retrieving web-based information on CHEM modules, datasets and analysis programs authored with the framework and for partially automating updates to a project documentation website.

@@

A library for **managing CHEM data** contains tools for supplying CHEM modules with input data ingested from local (i.e. a user's computer) or remote (online repositories) locations, labelling CHEM module datasets and exporting CHEM module data to online repositories. A library for **authoring reproducible analyses** contains tools to help write programs that apply CHEM modules to compatible datasets for the purpose of undertaking health economic analyses. These analysis programs can be either self-documenting (code is integrated with plain English explanations of what it does) or trigger the creation of separate documents (e.g. a scientific manuscript).

@@

We created a GitHub organisation (a collection of code repositories) where code (libraries, programs and sub-routines) that we author for and with the framework is stored and version controlled. We configured individual repositories in our GitHub organisation to use GitHub actions to implement continuous integration. By default, code libraries authored with our framework will use continuous integration to assess compliance with policies specified by the Comprehensive R Archive Network (CRAN) [@CRAN2022]. To track our code coverage, we linked our GitHub organisation to an account we established at codecov [@codecov_2022]. To facilitate the creation and hosting of documentation websites, we enabled GitHub Pages in each repository used for code library development.

>>>>

## R libraries 
A library in the R language will typically depend on multiple other R libraries, all of which potentially have different authors. As the number of third-party dependencies of an R library grows so does the fragility of that library (e.g., the library may cease to work as intended due to changes in one of its dependency libraries). To reduce the fragility of our framework we implemented it as multiple R libraries rather than one R library. In total we authored six novel R libraries to implement the ready4 framework, all of which have distinct purposes and dependencies (Table \@ref(tab:cpkgs)).

One framework library provides a **foundation for modular CHEM implementations**. In modular model implementations, modules need to be able to share inputs and outputs with each other and to be run as independent models [@barros2023empowering]. To achieve this, the foundation framework library defines a template CHEM module (using R’s S4 class system), which can be used to create other CHEM modules with a common set of inherited properties. One of these inherited properties is a novel syntax of 15 core commands that enable CHEM module algorithms to be consistently named. The foundation library also contains tools for retrieving web-based information on CHEM modules, datasets and analysis programs authored with the framework and tools for partially automating updates to a project documentation website.

@@

A library for **managing CHEM data** contains tools for supplying CHEM modules with input data ingested from local (i.e., a user's computer) or remote (online repositories) locations, labelling CHEM module datasets and exporting CHEM module data to online repositories. A library for **authoring reproducible analyses** contains tools to help write programs that apply CHEM modules to compatible datasets for the purpose of undertaking health economic analyses. These analysis programs can be either self-documenting (code is integrated with plain English explanations of what it does) or trigger the creation of separate documents (e.g., a scientific manuscript).

@@

We created a GitHub organisation (a collection of code repositories) where source code that we author is stored and version controlled. We configured individual repositories in our GitHub organisation to use GitHub actions to implement continuous integration. By default, code libraries authored with our framework will use continuous integration to assess compliance with policies specified by the Comprehensive R Archive Network (CRAN) [@CRAN2022]. To track our code coverage, we linked our GitHub organisation to an account we established at codecov [@codecov_2022]. To facilitate the creation and hosting of documentation websites, we enabled GitHub Pages in each repository used for code library development.
